---
title: "Student Performance Analysis Using R"
author: "Tiny Project - R Programming"
date: "`r format(Sys.Date(), '%d %B %Y')`"
output:
  html_document:
    toc: true
    toc_float: true
    number_sections: true
    theme: readable
---

# 1. Introduction

This Tiny Project analyses student performance using R. It covers data importing, preprocessing, cleaning, statistical exploration, visualization, interpretation, findings, and conclusion.

## Objectives

- Import student performance data.
- Clean and validate the data.
- Calculate descriptive statistics.
- Study relationships between academic variables.
- Create graphs and plots.
- Perform a simple regression analysis.
- Interpret the results.

# 2. Packages

```{r setup, message=FALSE, warning=FALSE}
packages <- c("tidyverse", "knitr")
new_packages <- packages[!(packages %in% rownames(installed.packages()))]
if (length(new_packages) > 0)
  install.packages(new_packages, repos = "https://cloud.r-project.org")

library(tidyverse)
library(knitr)
options(scipen = 999)
```

# 3. Database / Data Importing

```{r import-data, message=FALSE, warning=FALSE}
data_file <- "student_performance.csv"
students <- read.csv(data_file, stringsAsFactors = FALSE)
head(students)
```

# 4. Dataset Description

| Variable | Description |
|---|---|
| Student_ID | Unique student ID |
| Gender | Gender category |
| Study_Hours | Average study hours per day |
| Attendance | Attendance percentage |
| Previous_Marks | Previous academic marks |
| Assignment_Score | Assignment score |
| Sleep_Hours | Average sleep hours |
| Final_Marks | Final examination marks |

```{r dimensions}
cat("Rows:", nrow(students), "\n")
cat("Columns:", ncol(students), "\n")
str(students)
summary(students)
```

# 5. Data Preprocessing and Cleaning

## Missing values

```{r missing}
missing_values <- colSums(is.na(students))
kable(data.frame(Variable = names(missing_values),
                 Missing_Values = as.integer(missing_values)),
      caption = "Missing Values")
```

## Duplicate records

```{r duplicates}
cat("Duplicate rows:", sum(duplicated(students)))
```

## Data types

```{r factor}
students$Gender <- as.factor(students$Gender)
str(students)
```

## Range validation

```{r validation}
validation <- students %>%
  summarise(
    Min_Study_Hours = min(Study_Hours),
    Max_Study_Hours = max(Study_Hours),
    Min_Attendance = min(Attendance),
    Max_Attendance = max(Attendance),
    Min_Final_Marks = min(Final_Marks),
    Max_Final_Marks = max(Final_Marks)
  )
kable(validation, caption = "Range Validation")
```

# 6. Statistical Exploration

## Descriptive statistics

```{r descriptive}
stats <- students %>%
  summarise(
    Mean_Final = mean(Final_Marks),
    Median_Final = median(Final_Marks),
    SD_Final = sd(Final_Marks),
    Mean_Study = mean(Study_Hours),
    Mean_Attendance = mean(Attendance),
    Mean_Previous = mean(Previous_Marks),
    Mean_Assignment = mean(Assignment_Score)
  )
kable(round(stats, 2), caption = "Descriptive Statistics")
```

## Gender-wise analysis

```{r gender-analysis}
gender_summary <- students %>%
  group_by(Gender) %>%
  summarise(
    Students = n(),
    Average_Final_Marks = mean(Final_Marks),
    Average_Attendance = mean(Attendance),
    Average_Study_Hours = mean(Study_Hours)
  )
kable(round(gender_summary, 2), caption = "Gender-wise Performance")
```

## Correlation analysis

```{r correlation}
correlations <- students %>%
  summarise(
    Study_Hours = cor(Study_Hours, Final_Marks),
    Attendance = cor(Attendance, Final_Marks),
    Previous_Marks = cor(Previous_Marks, Final_Marks),
    Assignment_Score = cor(Assignment_Score, Final_Marks),
    Sleep_Hours = cor(Sleep_Hours, Final_Marks)
  )
kable(round(correlations, 3), caption = "Correlation with Final Marks")
```

# 7. Graph and Plot Creation

```{r histogram, fig.width=7, fig.height=5}
ggplot(students, aes(Final_Marks)) +
  geom_histogram(binwidth = 5, boundary = 0, color = "black") +
  labs(title = "Distribution of Final Marks",
       x = "Final Marks", y = "Number of Students") +
  theme_minimal()
```

```{r study-plot, fig.width=7, fig.height=5}
ggplot(students, aes(Study_Hours, Final_Marks)) +
  geom_point(size = 3) +
  geom_smooth(method = "lm", se = FALSE) +
  labs(title = "Study Hours vs Final Marks",
       x = "Study Hours per Day", y = "Final Marks") +
  theme_minimal()
```

```{r attendance-plot, fig.width=7, fig.height=5}
ggplot(students, aes(Attendance, Final_Marks)) +
  geom_point(size = 3) +
  geom_smooth(method = "lm", se = FALSE) +
  labs(title = "Attendance vs Final Marks",
       x = "Attendance (%)", y = "Final Marks") +
  theme_minimal()
```

```{r previous-plot, fig.width=7, fig.height=5}
ggplot(students, aes(Previous_Marks, Final_Marks)) +
  geom_point(size = 3) +
  geom_smooth(method = "lm", se = FALSE) +
  labs(title = "Previous Marks vs Final Marks",
       x = "Previous Marks", y = "Final Marks") +
  theme_minimal()
```

```{r assignment-plot, fig.width=7, fig.height=5}
ggplot(students, aes(Assignment_Score, Final_Marks)) +
  geom_point(size = 3) +
  geom_smooth(method = "lm", se = FALSE) +
  labs(title = "Assignment Score vs Final Marks",
       x = "Assignment Score", y = "Final Marks") +
  theme_minimal()
```

```{r boxplot, fig.width=7, fig.height=5}
ggplot(students, aes(Gender, Final_Marks)) +
  geom_boxplot() +
  labs(title = "Final Marks by Gender",
       x = "Gender", y = "Final Marks") +
  theme_minimal()
```

# 8. Statistical Model

```{r regression}
model <- lm(Final_Marks ~ Study_Hours, data = students)
summary(model)
regression_table <- as.data.frame(summary(model)$coefficients)
kable(round(regression_table, 4), caption = "Regression Results")
```

# 9. Findings

1. The dataset contains 30 student records and 8 variables.
2. Missing-value and duplicate checks were performed.
3. Descriptive statistics summarize student performance.
4. Correlation analysis measures linear associations with final marks.
5. Scatter plots provide visual evidence of relationships.
6. Regression provides a simple statistical model for the relationship between study hours and final marks.
7. Results describe this sample and should not be generalized to all students.

# 10. Conclusion

This project demonstrates a complete R Markdown workflow for student-performance analysis, including data importing, cleaning, preprocessing, statistical analysis, visualization, regression, interpretation, and reporting.

# 11. Limitations

- The dataset contains only 30 students.
- The data is intended for educational demonstration.
- Correlation does not establish causation.
- Several factors affecting performance are not included.
- The sample should not be treated as representative of all students.

# 12. Future Scope

Future work can include a larger real-world dataset, multiple regression, hypothesis testing, machine-learning prediction, and an interactive R Shiny dashboard.

# 13. Reproducibility Check

```{r reproducibility}
cat("Rows:", nrow(students), "\n")
cat("Columns:", ncol(students), "\n")
cat("Missing values:", sum(is.na(students)), "\n")
cat("Duplicate rows:", sum(duplicated(students)), "\n")
```

# 14. References

1. R Core Team. *R: A Language and Environment for Statistical Computing*.
2. Posit. *R Markdown Documentation*.
3. Wickham, H. et al. *Welcome to the Tidyverse*.
4. R documentation for `read.csv()`, `summary()`, `cor()`, `lm()`, and `ggplot2`.
